---
title: "07_BACH2_targets"
author: "Jana Braunger"
date: '2022-07-29'
output: BiocStyle::html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
BiocStyle::html_document 

dir <- "/Users/janabraunger/Documents/Studium/Master/Boston/Masterarbeit/"
figdir = paste0(dir, "Figures/BACH2/")
if(!dir.exists(figdir)) dir.create(figdir)
knitr::opts_chunk$set(dev = c("png", "pdf"), fig.path = figdir)

set.seed(202207)
```

```{r load_libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(magrittr)
library(pheatmap)
```

# Load data
## Comparison map 
The comparison map was calculated as "young + 2 * old" with young and old being the
binarized maps using the LAS algorithm. So, 0 means there is no interaction between
the loci in both celltypes, 1 only an interaction (as part of a LAS) in young fibroblasts,
2 only in old fibroblasts and 3 an interaction in both cells.

```{r}
comp_map <- read.delim(paste0(dir, "Data/difference_maps/comp_map_BACH2_targets.csv"), sep = ",") %>%
  column_to_rownames('chr1')
comp_map[1:4, 1:4]
```

## Louvain and hierarchical clusters
```{r}
louvain_clusters <- read.delim(paste0(dir, "Data/difference_maps/louvain_clusters_BACH2.csv"), sep = ",") %>%
  mutate(cluster = paste0("cluster_", cluster))

hierarchical_clusters <- read.delim(paste0(dir, "Data/difference_maps/hierarchical_clusters_BACH2.csv"), sep = ",") %>%
  mutate(cluster = paste0("cluster_", cluster))

head(hierarchical_clusters)
```


## Normalized RNAseq counts
```{r}
counts <- read.delim(paste0(dir, "Data/count_data.csv"), sep = ",")
mean_counts <- read.delim(paste0(dir, "Data/mean_counts.csv"), sep = ",") %>%
  group_by(gene, age_group) %>%
  summarize(counts = mean(counts))
head(mean_counts)
```


# Louvain clusters
## Number of nodes per specificity type and cluster
```{r}
# get number of young- and old-specific, as well as shared interactions per locus
edge_types = data.frame('locus' = rownames(comp_map))
edge_types$young_specific <- rowSums(sapply(comp_map, '==', 1))
edge_types$old_specific <- rowSums(sapply(comp_map, '==', 2))
edge_types$shared <- rowSums(sapply(comp_map, '==', 3))
edge_types$max_edges <- colnames(edge_types[1:nrow(edge_types),2:4])[max.col(edge_types[1:nrow(edge_types),2:4], 
                                                     ties.method = "random")]

# add number of edges per type to the cluster information
results <- louvain_clusters %>%
  left_join(edge_types) %>%
  mutate(max_edges = factor(max_edges, levels = c('young_specific', 'shared', 'old_specific')))
head(results)
```

```{r node_type_purity}
ggplot(results, aes(x = max_edges, fill = max_edges)) +
  geom_histogram(stat = "count") + 
  facet_wrap(~cluster, scale = "free_y", ncol = 3) +
  theme_bw() + 
  theme(axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.title.x = element_blank()) + 
  guides(fill=guide_legend(title="Node type")) +
  ylab('Number of nodes per category') +
  theme(text = element_text(size = 20))
```

## Average number of edges per type and cluster
```{r mean_edges}
mean_edges <- subset(results, select = c(locus, cluster,young_specific,
                                         old_specific, shared)) %>%
  gather(key = "type", value = value, -c(locus, cluster)) %>%
  group_by(cluster, type) %>%
  summarize(mean = mean(value)) %>%
  mutate(type = factor(type, levels = c('young_specific', 'shared', 'old_specific')))

ggplot(mean_edges, aes(x = type, y = mean, fill = type)) + 
  geom_bar(stat = 'identity') + 
  facet_wrap(~cluster, ncol = 3) +
  theme_bw() + 
  theme(axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.title.x = element_blank()) + 
  guides(fill=guide_legend(title="Edge type")) +
  ylab('Mean number of edges per locus') +
  theme(text = element_text(size = 20))
```

## Distribution of the degrees 
```{r degree_hist}
results %<>% mutate(degree = rowSums(subset(results, select = c(young_specific, old_specific, shared))))
ggplot(results, aes(x = degree, fill = cluster)) +
  geom_histogram() + 
  theme_bw() +
  facet_wrap(~cluster)
```


## Expression patterns
```{r expression_pattern}
# Add expression values
expr_targets <- inner_join(results, mean_counts) %>%
  group_by(gene) %>%
  mutate(counts = scale(counts)) %>%
  ungroup() %>%
  group_by(cluster, age_group) %>%
  slice_max(order_by = degree, n = 10)

ggplot(expr_targets, aes(x = age_group, y = counts, group = gene)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~cluster, nrow = 2)+
  geom_line(stat = "summary", fun = "median", colour = "brown", size = 1.2, aes(group = 1)) +
  theme_bw()+
  theme(text = element_text(size = 15))
```

# Hierarchical clustering
## Expression patterns for cluster 4
```{r expression_cluster4, fig.height = 3}
# Get expression values for genes in cluster 4
expr_targets <- inner_join(hierarchical_clusters, mean_counts) %>%
  filter(cluster == "cluster_4") %>%
  subset(select = c(gene, age_group, counts)) %>%
  spread(key = age_group, value = counts) 

# Create matrix
hclust_matrix <- select(expr_targets, -gene) %>%
  as.matrix()
rownames(hclust_matrix) <- expr_targets$gene

# Z-transform values for each gene
hclust_matrix %<>% t() %>% scale() %>% t()

# Calculate euclidean distances between genes
gene_dist <- dist(hclust_matrix)

# Clustering
gene_hclust <- hclust(gene_dist, method = "ward.D2")

# Plot dendrogram
plot(gene_hclust, labels = FALSE)

# Cut tree  
gene_cluster <- cutree(gene_hclust, k = 3) %>% 
  enframe() %>% 
  dplyr::rename(gene = name, cluster = value) %>%
  left_join(group_by(mean_counts, gene) %>% mutate(counts = scale(counts)))

# Plot expression patterns
ggplot(gene_cluster, aes(x = age_group, y = counts, group = gene)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~cluster, nrow = 1)+
  geom_line(stat = "summary", fun = "median", colour = "brown", size = 1.2, aes(group = 1)) +
  theme_bw()  +
  ylab('normalized RNAseq counts')+
  theme(text = element_text(size = 14))
```

## Expression patterns for cluster 5
```{r expression_cluster5, fig.height = 3}
# Get expression values for genes in cluster 5
expr_targets <- inner_join(hierarchical_clusters, mean_counts) %>%
  filter(cluster == "cluster_5") %>%
  subset(select = c(gene, age_group, counts)) %>%
  spread(key = age_group, value = counts) 

# Create matrix
hclust_matrix <- select(expr_targets, -gene) %>%
  as.matrix()
rownames(hclust_matrix) <- expr_targets$gene

# Z-transform values for each gene
hclust_matrix %<>% t() %>% scale() %>% t()

# Calculate euclidean distances between genes
gene_dist <- dist(hclust_matrix)

# Clustering
gene_hclust <- hclust(gene_dist, method = "ward.D2")

# Plot dendrogram
plot(gene_hclust, labels = FALSE)

# Cut tree  
gene_cluster <- cutree(gene_hclust, k = 3) %>% 
  enframe() %>% 
  dplyr::rename(gene = name, cluster = value) %>%
  left_join(group_by(mean_counts, gene) %>% mutate(counts = scale(counts)))

# Plot expression patterns
ggplot(gene_cluster, aes(x = age_group, y = counts, group = gene)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~cluster, nrow = 1)+
  geom_line(stat = "summary", fun = "median", colour = "brown", size = 1.2, aes(group = 1)) +
  theme_bw() +
  ylab('normalized RNAseq counts')+
  theme(text = element_text(size = 14))
```

## Expression patterns for all BACH2 targets
```{r expression_cluster_BACH2, fig.height = 3}
# Get expression values for targets
expr_targets <- inner_join(hierarchical_clusters, mean_counts) %>%
  subset(select = c(gene, age_group, counts)) %>%
  spread(key = age_group, value = counts) 

# Create matrix
hclust_matrix <- select(expr_targets, -gene) %>%
  as.matrix()
rownames(hclust_matrix) <- expr_targets$gene

# Z-transform values for each gene
hclust_matrix %<>% t() %>% scale() %>% t()

# Calculate euclidean distances between genes
gene_dist <- dist(hclust_matrix)

# Clustering
gene_hclust <- hclust(gene_dist, method = "ward.D2")

# Plot dendrogram
plot(gene_hclust, labels = FALSE)

# Cut tree  
gene_cluster <- cutree(gene_hclust, k = 6) %>% 
  enframe() %>% 
  dplyr::rename(gene = name, cluster = value) %>%
  left_join(group_by(mean_counts, gene) %>% mutate(counts = scale(counts)))

# Plot expression patterns
ggplot(gene_cluster, aes(x = age_group, y = counts, group = gene)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~cluster, nrow = 2)+
  geom_line(stat = "summary", fun = "median", colour = "brown", size = 1.2, aes(group = 1)) +
  theme_bw() +
  ylab('normalized RNAseq counts')+
  theme(text = element_text(size = 14))

write.csv(gene_cluster, paste0(dir, "Data/BACH2_expression_clusters.csv"), row.names = FALSE)
```








## Expression patterns for all expressed genes



