---
title: "TF_clustering"
author: "Jana"
date: "2023-05-04"
output: BiocStyle::html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
BiocStyle::html_document 

dir <- "/Users/janabraunger/Documents/Studium/Master/Boston/Masterarbeit/"
figdir = paste0(dir, "Figures/TF_clustering/")
if(!dir.exists(figdir)) dir.create(figdir)
knitr::opts_chunk$set(dev = c("png", "pdf"), fig.path = figdir)

set.seed(202210)
library(showtext)
font_add("Arial", "/Library/Fonts/Arial.ttf")
showtext_auto()
```

```{r load_libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(magrittr)
library(pheatmap)
library(RColorBrewer)
```

This markdown includes the code to create the clustered heatmaps in Figure 5C, as well as the boxplots in Figure 5A and B.

# Load data
## TF centrality and TF enrichment for targeting DE genes
```{r}
TF_centrality_young <- read.delim(paste0(dir, "Data/pcst/TFs_centrality_young.csv"), sep = ",") %>%
  relocate(TF) %>%
  mutate(adj_pval = ifelse(adj_pval > 1, 1, adj_pval))
TF_centrality_old <- read.delim(paste0(dir, "Data/pcst/TFs_centrality_old.csv"), sep = ",") %>%
  relocate(TF) %>%
  mutate(adj_pval = ifelse(adj_pval > 1, 1, adj_pval))
tail(TF_centrality_old)
```


## Get groups of shared, young-, old-specific TFs
```{r}
pcst_groups <- read.delim(paste0(dir, "Data/specific_TFs_targets.csv"), sep = ",") %>%
  subset(select = c('TF', 'net')) %>%
  distinct()
head(pcst_groups)
```


## TF expression
```{r}
counts <- read.delim(paste0(dir, "Data/count_data.csv"), sep = ",") %>%
  subset(select = c(gene, age_group, counts)) %>%
  group_by(gene, age_group) %>%
  summarize(counts = mean(counts)) %>% 
  ungroup() %>%
  group_by(gene) %>%
  summarize(mean_expr = mean(counts), var_expr = var(counts)) %>%
  dplyr::rename(TF = gene)
head(counts)
```


## Intermingling between DE targets per TF
```{r}
int_young_DE <- read.delim(paste0(dir, "Data/spec_intermingling_young.csv"), sep = ",")
int_old_DE <- read.delim(paste0(dir, "Data/spec_intermingling_old.csv"), sep = ",")
head(int_young_DE)
```


# Clustered heatmaps
## Steiner network S1
a) Including all bridge TFs 
```{r young_net_selected_cols, fig.height=10, fig.width=3}
TF_features <- TF_centrality_young %>%
  subset(select = -net) %>%
  left_join(counts) %>%
  left_join(int_young_DE) %>%
  mutate(DE_enrichment = adj_pval) %>%
  gather(key = 'measure', value = 'value', -TF) %>%
  group_by(measure) %>%
  mutate(value = scale(value)) %>%
  mutate(value = ifelse(value < -2, -2, value)) %>%
  mutate(value = ifelse(value > 2, 2, value)) %>%
  spread(key = measure, value = value) %>%
  column_to_rownames(var = 'TF') %>%
  subset(select = c('DE_enrichment', 'spec_intermingling')) %>%
  arrange(DE_enrichment, desc(spec_intermingling))

anno <- filter(pcst_groups, net %in% c('shared')) %>%
  subset(select = c('TF', 'net')) %>% distinct() %>%
  mutate(TF = factor(TF, levels = rownames(TF_features))) %>%
  arrange(TF) %>%
  column_to_rownames(var = 'TF')
colnames(anno) <- c('Shared Bridge TFs \n (in S1, S2 and S3)')
breakslist = seq(-2, 2, by = 0.1)

pheatmap(TF_features, cluster_cols = FALSE, cluster_rows = FALSE, 
         annotation_row = anno, annotation_legend = FALSE, 
         labels_col = c('Enrichment p-value of \n DE target genes in S1',
                        '% of specific intermingling \n between the DE target \n genes in S1'), 
         fontsize = 11, fontsize_col = 14, fontsize_row = 10, breaks = breakslist, 
         color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breakslist)))
```

b) Clustering for significant bridge TFs
```{r young_net_clustering, fig.height=8}
signif_TFs <- c('CEBPB', 'FLI1', 'MAZ', 'SUMO2', 'AHR', 'EP300')
TF_features <- TF_centrality_young %>%
  subset(select = -net) %>%
  filter(TF %in% signif_TFs) %>%
  left_join(counts) %>%
  left_join(int_young_DE) %>%
  mutate(DE_enrichment = adj_pval) %>%
  gather(key = 'measure', value = 'value', -TF) %>%
  group_by(measure) %>%
  #mutate(value = rank(value, ties.method = 'average')) %>%
  #mutate(value = (value - min(value)) / (max(value) - min(value))) %>%
  mutate(value = scale(value)) %>%
  mutate(value = ifelse(value < -2, -2, value)) %>%
  mutate(value = ifelse(value > 2, 2, value)) %>%
  spread(key = measure, value = value) %>%
  column_to_rownames(var = 'TF') %>%
  subset(select = c('mean_expr', 'var_expr', 'all_targets', 'DE_targets', 'katz_centrality', 
                    'PPIs', 'DE_enrichment', 'spec_intermingling', 'shared_intermingling'))

anno <- filter(pcst_groups, net %in% c('shared')) %>%
  filter(TF %in% signif_TFs) %>%
  subset(select = c('TF', 'net')) %>% distinct() %>%
  mutate(TF = factor(TF, levels = rownames(TF_features))) %>%
  arrange(TF) %>%
  column_to_rownames(var = 'TF')
colnames(anno) <- c('Shared Bridge TFs \n (in S1, S2 and S3)')

out <- pheatmap(subset(TF_features, select = c('DE_targets', 'katz_centrality', 'PPIs', 
                                        'DE_enrichment', 'spec_intermingling')), 
                cluster_cols = FALSE, annotation_row = anno)
TF_order <- rownames(TF_features[out$tree_row[["order"]], ])
```


```{r young_net, fig.height=6, fig.width=10}
pheatmap(TF_features[TF_order,], cluster_cols = FALSE, cluster_rows = FALSE, 
         annotation_row = anno, annotation_legend = FALSE, 
         labels_col = c('Mean of the median \n expression per age group', 'Variance of the median \n expression per age group',
                        'Number of targets \n in the genome', 'Number of targets in \n the S1 target DE genes',
                        'Katz centrality \n in S1', 'Number of PPI \n neighbors in S1', 
                        'Enrichment p-value of \n DE target genes in S1',
                        '% of specific intermingling \n between the DE target \n genes in S1', 
                        '% of consistent intermingling \n between the DE target \n genes in S1'), 
         fontsize = 11, fontsize_col = 14, fontsize_row = 10)
```

## Steiner network S3
a) For all bridge TFs
```{r old_net_selected_cols, fig.height=10, fig.width = 3}
TF_features <- TF_centrality_old %>%
  subset(select = -net) %>%
  left_join(counts) %>%
  left_join(int_old_DE) %>%
  mutate(DE_enrichment = adj_pval) %>%
  gather(key = 'measure', value = 'value', -TF) %>%
  group_by(measure) %>%
  mutate(value = scale(value)) %>%
  mutate(value = ifelse(value < -2, -2, value)) %>%
  mutate(value = ifelse(value > 2, 2, value)) %>%
  spread(key = measure, value = value) %>%
  column_to_rownames(var = 'TF') %>%
  subset(select = c('DE_enrichment', 'spec_intermingling')) %>%
  arrange(DE_enrichment, desc(spec_intermingling))

anno <- filter(pcst_groups, net %in% c('shared')) %>%
  subset(select = c('TF', 'net')) %>% distinct() %>%
  mutate(TF = factor(TF, levels = rownames(TF_features))) %>%
  arrange(TF) %>%
  column_to_rownames(var = 'TF')
colnames(anno) <- c('Shared Bridge TFs \n (in S1, S2 and S3)')

breakslist = seq(-2, 2, by = 0.1)
pheatmap(TF_features[, c('DE_enrichment', 'spec_intermingling')], cluster_cols = FALSE, cluster_rows = FALSE, 
         annotation_row = anno, annotation_legend = FALSE, 
         labels_col = c('Enrichment p-value of \n DE target genes in S3',
                        '% of specific intermingling \n between the DE target \n genes in S3'), 
         fontsize = 11, fontsize_col = 14, fontsize_row = 10, breaks = breakslist, 
         color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breakslist)))
```

b) For significant bridge TFs
```{r old_net_clustering, fig.height=8}
signif_TFs <- c('E2F1', 'KDM4C', 'POLR2A', 'HOXA6', 'HDAC1', 'TAF3', 'GTF2B',
                'RBBP5', 'NOTCH1', 'GABPA', 'MNT', 'BRD2', 'KAT5', 'BMI1', 'ZNF384')
TF_features <- TF_centrality_old %>%
  subset(select = -net) %>%
  filter(TF %in% signif_TFs) %>%
  left_join(counts) %>%
  left_join(int_old_DE) %>%
  mutate(DE_enrichment = adj_pval) %>%
  gather(key = 'measure', value = 'value', -TF) %>%
  group_by(measure) %>%
  mutate(value = scale(value)) %>%
  mutate(value = ifelse(value < -2, -2, value)) %>%
  mutate(value = ifelse(value > 2, 2, value)) %>%
  spread(key = measure, value = value) %>%
  column_to_rownames(var = 'TF') %>%
  subset(select = c('mean_expr', 'var_expr', 'all_targets', 'DE_targets', 'katz_centrality',
                    'PPIs', 'DE_enrichment', 'spec_intermingling', 'shared_intermingling'))

anno <- filter(pcst_groups, net %in% c('shared')) %>%
  filter(TF %in% signif_TFs) %>%
  subset(select = c('TF', 'net')) %>% distinct() %>%
  mutate(TF = factor(TF, levels = rownames(TF_features))) %>%
  arrange(TF) %>%
  column_to_rownames(var = 'TF')
colnames(anno) <- c('Shared Bridge TFs \n (in S1, S2 and S3)')

out <- pheatmap(subset(TF_features, select = c('DE_targets', 'katz_centrality', 'PPIs', 
                                        'DE_enrichment', 'spec_intermingling')), 
                cluster_cols = FALSE, annotation_row = anno)
TF_order <- rownames(TF_features[out$tree_row[["order"]], ])
```

```{r old_net, fig.height=9, fig.width = 10}
pheatmap(TF_features[TF_order,], cluster_cols = FALSE, cluster_rows = FALSE, 
         annotation_row = anno, annotation_legend = FALSE, 
         labels_col = c('Mean over the mean \n expression per age group', 'Variance of the mean \n expression per age group',
                        'Number of targets \n in the genome', 'Number of targets in \n the S3 target DE genes',
                        'Katz centrality \n in S3', 'Number of PPI \n neighbors in S3', 
                        'Enrichment p-value of \n DE target genes in S3',
                        '% of specific intermingling \n between the DE target \n genes in S3', 
                        '% of consistent intermingling \n between the DE target \n genes in S3'), 
         fontsize = 11, fontsize_col = 14, fontsize_row = 10)
```


# More cell state specific intermingling for significant TFs?
```{r intermingling_signif_TFs, fig.width=10}
intermingling_signif <- TF_centrality_young %>%
  bind_rows(TF_centrality_old) %>%
  subset(select = c(TF, net, adj_pval)) %>%
  mutate(significant_TFs = ifelse(adj_pval < 0.05, 'Significant \n DE targeting', 'Non significant \n DE targeting')) %>%
  left_join(bind_rows(int_young_DE %>% mutate(net = "young"), int_old_DE %>% mutate(net = "old"))) %>%
  mutate(net = ifelse(net == 'young', 'Steiner network S1', 'Steiner network S3')) %>%
  mutate(net = factor(net, levels = c('Steiner network S1', 'Steiner network S3')))

ggplot(intermingling_signif, aes(x = significant_TFs, y = spec_intermingling, fill = significant_TFs)) + 
  geom_boxplot(outlier.size = 0) + 
  geom_jitter(width = 0.15, height = 0, alpha = 0.4, size = 2) +
  facet_wrap(~net) + 
  theme_bw() +
  ylab('% of specific intermingling between \n the target DE genes per TF') + 
  xlab('Bridge TFs') +
  theme(text = element_text(size = 18)) + 
  scale_fill_manual(values=c("#aaaaaa", "#c6d02a"))
```


```{r}
# T-test for young
t.test(dplyr::filter(intermingling_signif, net == "Steiner network S1", 
                     significant_TFs == 'Significant \n DE targeting')$spec_intermingling, 
       dplyr::filter(intermingling_signif, net == "Steiner network S1", 
                     significant_TFs == 'Non significant \n DE targeting')$spec_intermingling, 
       mu = 0, alternative = "greater")

# T-test for old
t.test(dplyr::filter(intermingling_signif, net == "Steiner network S3", 
                     significant_TFs == 'Significant \n DE targeting')$spec_intermingling, 
       dplyr::filter(intermingling_signif, net == "Steiner network S3", 
                     significant_TFs == 'Non significant \n DE targeting')$spec_intermingling, 
       mu = 0, alternative = "greater")
```


```{r intermingling_non_targeted_DE_genes, fig.width=18}
# Instead of a boxplot, make a barplot and add DE genes targeted by no bridge TFs
im_barplot <- intermingling_signif %>% 
  bind_rows(data.frame(TF = 'None', net = 'Steiner network S1', adj_pval = 1, 
                       significant_TFs = 'DE genes not\ntargeted by any\nbridge TF', 
                       spec_intermingling = 0.0078125, shared_intermingling = 0)) %>%
  bind_rows(data.frame(TF = 'None', net = 'Steiner network S3', adj_pval = 1, 
                       significant_TFs = 'DE genes not\ntargeted by any\nbridge TF', 
                       spec_intermingling = 0.008888888888888889, shared_intermingling = 0))
im_barplot$significant_TFs[im_barplot$significant_TFs == 'Non significant \n DE targeting'] <- 'DE genes targeted\nby bridge TFs with non\nsignificant DE targeting'
im_barplot$significant_TFs[im_barplot$significant_TFs == 'Significant \n DE targeting'] <- 'DE genes targeted\nby bridge TFs with\nsignificant DE targeting'
im_barplot %<>% mutate(significant_TFs = factor(significant_TFs, 
                                                levels = c('DE genes not\ntargeted by any\nbridge TF',
                                                           'DE genes targeted\nby bridge TFs with non\nsignificant DE targeting',
                                                           'DE genes targeted\nby bridge TFs with\nsignificant DE targeting'
                                                           )))

medians <- im_barplot %>%
  group_by(net, significant_TFs) %>%
  summarise(median_value = median(spec_intermingling, na.rm = TRUE))

ggplot(im_barplot, aes(x = significant_TFs, y = spec_intermingling, fill = significant_TFs)) + 
  geom_bar(data = medians, aes(x = significant_TFs, y = median_value), stat = "identity", width = 0.75) +
  #geom_jitter(width = 0.15, height = 0, alpha = 0.4, size = 2) +
  facet_wrap(~net) + 
  theme_bw() +
  ylab('% of specific intermingling\nbetween selected DE genes') + 
  xlab('') +
  theme(text = element_text(size = 18)) + 
  scale_fill_manual(values=c("#848484", "#aaaaaa", "#c6d02a"))
```

```{r intermingling_non_DE, fig.width=10}
# Intermingling of non-DE genes
int_young_DE <- read.delim(paste0(dir, "Data/spec_intermingling_random_young.csv"), sep = ",")
int_old_DE <- read.delim(paste0(dir, "Data/spec_intermingling_random_old.csv"), sep = ",")

intermingling_signif <- TF_centrality_young %>%
  bind_rows(TF_centrality_old) %>%
  subset(select = c(TF, net, adj_pval)) %>%
  mutate(significant_TFs = ifelse(adj_pval < 0.05, 'Significant \n DE targeting', 'Non significant \n DE targeting')) %>%
  left_join(bind_rows(int_young_DE %>% mutate(net = "young"), int_old_DE %>% mutate(net = "old"))) %>%
  mutate(net = ifelse(net == 'young', 'Steiner network S1', 'Steiner network S3')) %>%
  mutate(net = factor(net, levels = c('Steiner network S1', 'Steiner network S3')))

ggplot(intermingling_signif, aes(x = significant_TFs, y = spec_intermingling, fill = significant_TFs)) + 
  geom_boxplot(outlier.size = 0) + 
  geom_jitter(width = 0.15, height = 0, alpha = 0.4, size = 2) +
  facet_wrap(~net) + 
  theme_bw() +
  ylab('% of specific intermingling between \n the non-DE target genes per TF') + 
  xlab('Bridge TFs') +
  theme(text = element_text(size = 18)) + 
  scale_fill_manual(values=c("#aaaaaa", "#c6d02a"))
```

```{r}
# T-test for young
t.test(dplyr::filter(intermingling_signif, net == "Steiner network S1", 
                     significant_TFs == 'Significant \n DE targeting')$spec_intermingling, 
       dplyr::filter(intermingling_signif, net == "Steiner network S1", 
                     significant_TFs == 'Non significant \n DE targeting')$spec_intermingling, 
       mu = 0, alternative = "greater")

# T-test for old
t.test(dplyr::filter(intermingling_signif, net == "Steiner network S3", 
                     significant_TFs == 'Significant \n DE targeting')$spec_intermingling, 
       dplyr::filter(intermingling_signif, net == "Steiner network S3", 
                     significant_TFs == 'Non significant \n DE targeting')$spec_intermingling, 
       mu = 0, alternative = "greater")
```


# Number of TFs targeting each DE gene
```{r n_TFs_per_gene, fig.width = 10}
target_counts_young <- read.delim(paste0(dir, "Data/n_TFs_per_young_DE_gene.csv"), sep = ",") %>%
  mutate(net = "Steiner network S1")
target_counts_old <- read.delim(paste0(dir, "Data/n_TFs_per_old_DE_gene.csv"), sep = ",") %>%
  mutate(net = "Steiner network S3")
target_counts <- bind_rows(target_counts_young, target_counts_old) %>%
  mutate(net = factor(net, levels = c('Steiner network S1', 'Steiner network S3'))) %>%
  mutate(intermingling = recode(intermingling, True = "With specific \n intermingling", False = "No specific \n intermingling"))

ggplot(target_counts, aes(x = intermingling, y = n_TFs, fill = intermingling)) + 
  geom_boxplot() + 
  geom_jitter(width = 0.15, height = 0, alpha = 0.4, size = 2) +
  facet_wrap(~net) + 
  theme_bw() +
  ylab('Number of Bridge TF regulators') + 
  xlab('Target DE genes') +
  theme(text = element_text(size = 18)) + 
  scale_fill_manual(values=c("#aaaaaa", "#c6d02a"))
```


```{r}
# T-test for young
t.test(dplyr::filter(target_counts, net == "Steiner network S1", intermingling == 'With specific \n intermingling')$n_TFs, 
       dplyr::filter(target_counts, net == "Steiner network S1", intermingling == "No specific \n intermingling")$n_TFs, 
       mu = 0, alternative = "greater")

# T-test for old
t.test(dplyr::filter(target_counts, net == "Steiner network S3", intermingling == "With specific \n intermingling")$n_TFs, 
       dplyr::filter(target_counts, net == "Steiner network S3", intermingling == "No specific \n intermingling")$n_TFs, 
       mu = 0, alternative = "greater")
```
